---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  fig.path = "tools/"
)
```

# latte

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/latte)](https://cran.r-project.org/package=latte)
[![Travis build status](https://travis-ci.org/dkahle/latte.svg?branch=master)](https://travis-ci.org/dkahle/latte)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/dkahle/latte?branch=master&svg=true)](https://ci.appveyor.com/project/dkahle/latte)
<!-- badges: end -->

__latte__ is an R package that makes back-end connections to [LattE](https://www.math.ucdavis.edu/~latte/software.php) and [4ti2](http://www.4ti2.de).  Most of its functions were previously part of the [__algstat__ package](https://github.com/dkahle/algstat), but have been pulled out and improved upon.

It is currently being developed.


_Note: the following assumes you have [LattE](https://www.math.ucdavis.edu/~latte/) and [4ti2](http://www.4ti2.de) installed and latte recognizes their path._

```{r load}
library("latte")
```







## Lattice point counting

Most [LattE](https://www.math.ucdavis.edu/~latte/) programs are available as functions in __latte__.  For example, `latte_count()` uses LattE's `count` to determine the number of integer points in a [polytope](http://en.wikipedia.org/wiki/Polytope):

```{r latte-count}
latte_count(c("x + y <= 10", "x >= 0", "y >= 0"))
```

It's easy to confirm the solution with a simple visualization:

```{r countExample, fig.height=4, dpi=200}
library(ggplot2); theme_set(theme_bw())
polytope <- data.frame(x = c(0,10,0), y = c(0,0,10))
points   <- expand.grid(x = 0:10, y = 0:10)
points   <- subset(points, x + y <= 10)
points$number <- 1:nrow(points)
ggplot(aes(x = x, y = y), data = polytope) +
  geom_polygon(fill = "red", alpha = .2) + 
  geom_text(aes(y = y + .25, label = number), size = 3.5, data = points) +
  geom_point(data = points) + 
  coord_equal()
```



## Integer programming

In addition to table counting, it can also do integer programming with LattE's `latte-maximize` and `latte-minimize` programs.  To do this, it uses tools from [__mpoly__](http://github.com/dkahle/mpoly):

```{r ip}
latte_max("-2 x + 3 y", c("x + y <= 10", "x >= 0", "y >= 0"))
latte_min("-2 x + 3 y", c("x + y <= 10", "x >= 0", "y >= 0"))
```

We can check that the solution given above is correct, but the value is not.  So, it needs some more work:

```{r ipCheck, fig.height=4, dpi=200}
points$objective <- with(points, -2*x + 3*y)
ggplot(aes(x = x, y = y), data = polytope) +
  geom_polygon(fill = "red", alpha = .2) + 
  geom_point(aes(size = objective), data = points) + 
  coord_equal()
```








## Installation

* From Github (dev version): 
```{r, eval=FALSE}
if (!requireNamespace("devtools")) install.packages("devtools")
devtools::install_github("dkahle/mpoly")
devtools::install_github("dkahle/latte")
```





## Acknowledgements

This material is based upon work supported by the National Science Foundation under Grant Nos. [1622449](https://nsf.gov/awardsearch/showAward?AWD_ID=1622449) and [1622369](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1622369).